Modelling Prehistoric Iconographic Compositions. The \proglang{R} package \pkg{decorr}

### 2020-12-12 article_rvTH16.Rnw
up

Symbolic practices is a characteristic trait of human societies. Even discussed, such practices seems to start between 233,000 to 800,00 BC \citep{dErrico00}, covering more than 97\% of total human societies time span. Symbolism cover a large range of practices, from ochre deposit in a tomb, to menhir alignements, passing through wall fresco. This latter, what might be called "iconographical practices", probably shows the most complex and interesting testimonies of past societies symbolism. For decades, its study was linked to history of religion because commonly seen as closely linked to cultual practices and believes. Since the \textit{New Archaeology} developpement during the 60's \citep{Clarke14}, symbolic expressions start to be studied with the same formal methods (statistics, seriations, distribution maps, etc.) as any another aspect of social organisation: settlement patterns, tools \emph{chaine op\'eratoire}, susbsitence strategies, etc. \citep{Renfrew91}. But unlike many aspects of the material culture where technological requirements and technical efficiency determine the choice of the raw material and the shape of the object -- a flint blade for cutting, a pottery for containing, a house for living --, the function of an iconographic composition cannot be drawn directly from itself. Whether these last decades study of ancient iconography had  undergone significative improvements at the site scale -- with GIS/database statistics, paleoenvironmental restitutions, etc. -- and at the sign scale with the development of archaeological sciences -- radiocarbon dating, use-wear analysis, elemental analysis, etc. --, these improvement do not necessarly help to understand the semantic content of the iconography.\\
Semantics or semiotics can be defined as a system of conventional features -- called signs -- organised also in conventional manners. Until our days, formal methods to study ancient iconography semantics have been mostlty been grounded  -- explicitly or not -- on the prime principle of Saussurian linguistic: the '\emph{linearity of the signifier}' \citep{Saussure89}.
Writing is one of the most rational semiographical system with a clear distinction between signified and signifier and the development of the signified on a linear axis. Even if we do not understand the meaning of the signs, writing can easily be modelled with Graph theory and recurent patterns can be identified. For example, the 3-letters word "\code{art}" can be modelled with three vertices (\code{a}, \code{r}, \code{t}) and two edges (one between \code{a} and \code{r}, the other between \code{r} and \code{t}). In \proglang{R}, these features, concatenated in this order with a \code{paste0()}, is \code{art}, and not \code{rat} (Fig. \ref{fig:fart}).

@@

Because of the inherent variability of iconography, and because graphical and spatial proximities between graphical units are generaly not quantified, applying the Saussurian model to any prehistorical graphical content had led to considerable problems:
\begin{itemize}
\setlength\itemsep{.05em}
\item unexplicit groupings of graphical units -- like graphical units grouped into \emph{figures},  \emph{figures} grouped into \emph{patterns}, \emph{patterns} grouped into \emph{motives}, etc. -- with tedious number of groups
\item consistency, proximities and relationships between these groups are often implicit and not quantified
\item studies develop proper descriptive vocabularies, singular relationships of grouping, idosyncratic methods at site-dependent or period-dependent scales
\end{itemize}

@@

According to us, prehistorical composition -- like any formal system -- are spatial features (mostly shapes) related one with the other depending on rules of spatial proximities. Then, the First Law of Geography: "\emph{everything is related to everything else, but near things are more related than distant things}" \citep{Tobler70} should be more operative than the Saussurian '\emph{linearity of the signifier}' principle \citep{Saussure89}. To get these '\emph{near things}', or features proximities, we use concepts coming from the Graph Theory and particularly relative neighborhood graphs (RNG): '\emph{a powerful model of low-level visual process}' \citep{Toussaint80}.

@@

Nodes and edges -- repectively GUs and connexions between GUs -- are created on a GIS interface. Indeed, for large series of decorated supports, GIS offers the most suitable and flexible interface to register all GUs and to get their coordinates. The decoration image is considerated as the basemap of the project and will cover the region of interest of the analysis (ie, all the graphical content). The decoration image can be binarized: GUs are considerated active, the undecorated parts of the support -- the background -- are considerated inactive. The polygonisation of the GUs (\code{POLYGONS}), including the border of the stelae, allows to calculated their Voronoi seeds (ie, their centroid) and Voronoi cells (ie, their area of influence). So, the entire decoration image is tiled and condidered as a Voronoi diagram. Centroids of the GUs (\code{POINTS}) are calculated and considered as the graph nodes, their locations (\emph{x} and \emph{y} coordinates, measured in pixels) are relative to each decoration. Exist edges (\code{LINES}) between pairs of GUs when their Voronoi cells share a common border (\emph{birel}: 'touches').

The entire graphical content of the decoration is mapped with a neighborhood graphs (RNG) -- ie geometric graph, or planar graph -- of GUs centroids (nodes) and GUs proximity links (edges). So a decoration graph is a one component graph which is the dual graph of the Voronoi diagram (Fig. \ref{fig:gis}).

\begin{figure}[!htbp]
\centering
\includegraphics{gis}
\caption{\label{fig:gis} GIS interface. A) Original decoration of the Late Bronze Age \emph{Cerro Muriano 1} stelae (drawing: \cite{DiazGuardamino10}) opened in new GIS project without any projection system with its extent (\code{xmin, xmax, ymin, ymax} in pixels); B) Polygonsiation of the GUs, calculation of their centroids (nodes), calculation of the Voronoi cells, calculation of the proximity links (edges); C) plot of the relative \emph{x} and \emph{y} coordinates of the GUs centroids; D) plot of the values in the \code{type} of the GUs}
\end{figure}

Whether this theoritical process include image binarization, polygonization and \emph{birel} analysis, a simplier solution will be to create directly centroids on the GUs and to draw the edges manually. Then \emph{x} and \emph{y} coordinates of nodes can be easily retrieved.\\
This model has a minimal of \emph{a priori} definitions. Those definitions only concern the intrinsic variables of GUs (\code{type}, \code{technologie}, \code{color}, \code{size}, etc.) and the types of relations they share. At least one variable, like the \code{type} of the GUs ('anthropomorph', 'horned helmet', etc.), is defined in order to compute composition analysis. Here, we will only consider one property for the nodes, its \code{type}, and the most common types of relations between GUs:

\begin{description}
\item[Normal edges]
By convention, two contemporaneous GUs having a Voronoi cell sharing a border, have a common edge tagged \code{'='}. This proximity is represented with a plain line. The textual notation of such an edge is \code{'-=-'}. For example: \code{personnage -=- bouclier} means that the nodes \code{personnage} and \code{bouclier} have a common border (Fig. \ref{fig:gis_1}).

\item[Attribute edges]
It occurs frequently that a GU can be divided into a \emph{main unit} (eg, an anthropomorph) and one or various \emph{attribute units} (eg, a helmet, a male sex). Broadly speaking, for further statistical analysis, it is better to use this \emph{attribute} method than to multiply the categories of GUs. To record this information, a new type of edge, tagged with \code{'+'}, is be introduced. This type of edges is directed and, by convention, displayed with a dashed line. Its starts from the \emph{main unit} and ends with the \emph{attribute unit}. The textual notation of such an edge is \code{'-+-'}. For example \code{personnage -+- casque} means that the main node \code{personnage} has the attribute node \code{casque} (Fig. \ref{fig:gis_1}).

\begin{figure}[!htbp]
\centering
\includegraphics[height=10cm, trim=15mm 20mm 15mm 20mm, clip, keepaspectratio]{gis_1}
\caption{\label{fig:gis_1} GIS interface, Cerro Muriano 1 decoration. The GUs \code{casque} (helmet) and \code{sexe\_masculin} (male sex) are two nodes attributes of the GU \code{personnage} (anthropomorph).}
\end{figure}

\item[Overlapping edges]
Finally, it is quite common that a graphical composition shows superimpositions between different GUs permitting to distinguish different decoration phases, or iconographic layers, for a single support. So, at first, the analyse must be performed on each different phases of decoration separately. This stratigraphical information (\code{A} \emph{over} \code{B}, or \code{B} \emph{under} \code{A}) helps to understand the relative chronology between GUs and must be recorded. A simple way to achieve this is to introduce the new tag \code{'>'} for the for the type of edge. This type of edges is directed. The textual notation of such an edge is \code{'->-'}. For example \code{ecriture ->- bouclier} means that \code{ecriture} crosses \code{bouclier} (ie, \code{ecriture} overlaps \code{bouclier} in the stratigraphical sense) (Fig. \ref{fig:gis_2}).

@@

\begin{figure}[!htbp]
\centering
\includegraphics[height=10cm, trim=30mm 16mm 30mm 13mm, clip, keepaspectratio]{gis_2}
\caption{\label{fig:gis_2} GIS interface, Ibahernando decoration. The GUs \code{ecriture} (writing) overlaps \code{lance} (spear) and the upper part of the \code{bouclier} (shield).}
\end{figure}

@@

\begin{table}[!htbp]
\centering
\begin{tabular}{cccccp{4.9cm}}
\hline
% node A & type of edge & node B & undirected & birel & stratigraphical meaning \\
GU.1 & Edge type & GU.2 & Edge directness & \emph{Birel} & Description \rule{0pt}{1em} \\[1pt]
\hline \rule{0pt}{1em}%
\code{A} & \code{=} & \code{B} & Undirected & $ A \cap B = \emptyset$ & 
  \code{A} and \code{B} are contemporaneous and disjoint. \\
\code{A} & \code{+} & \code{B} & Directed & $ A \cap B = A $ & 
  \code{A} and \code{B} are contemporaneous. \code{B} is an attribute of \code{A}. \\
\code{A} & \code{>} & \code{B} & Directed & $ A \cap B = \exists $ & 
  \code{A} is more recent than \code{B}.\newline \code{A} overlaps \code{B}. \\
% \code{1} & = & 2 & undirected & A \cap B = \varnothing & 1 and 2 can be contemporaneous \\
\hline
\end{tabular}
\caption{Synthesis for the different types of relations between GUs.}\label{tab1}
\label{table:tedges}
\end{table}

@@

A training dataset (nodes and edges coordinates, decoration images) is stored in the \code{extdata} folder of the \pkg{decorr} package
\smallbreak
% \begin{itemize}
% \item The \code{imgs} dataframe
% \end{itemize}
\textbf{\code{imgs} dataframe}
\\
The inventory of decorations is stored in the \code{imgs} dataframe (Tab. \ref{tab:timg}). The field \code{imgs$idf} is the short name of the decoration, useful during statistical analysis. The primary key of each decoration is the concatenate of \code{imgs$site} and \code{imgs$decor}. These keys will allow joints with the other dataframes (\code{nodes} and \code{edges})
%
<<results='asis', echo=FALSE, warning=FALSE>>=
# library("decorr")
# library("igraph")
# library("magick")
# # par(mar=c(0,0,0,0))
imgs <- read.table(system.file("extdata", "imgs.tsv", package = "decorr"),
                   sep="\t", stringsAsFactors = FALSE)
print(xtable::xtable(imgs,
    caption="The studied corpus, the \\code{imgs.tsv} dataframe",
    label="tab:timg",
    size=6,
    align=c(
      "p{0.01cm}",
      "c",
      ">{\\raggedright\\arraybackslash\\rule{0pt}{1em}\\addtolength{\\baselineskip}{-1.5pt}}p{2.1cm}",
      ">{\\raggedright\\arraybackslash\\addtolength{\\baselineskip}{-1.5pt}}p{2.1cm}",
      "p{8.5cm}")),
    table.placement="!htbp",
    include.rownames=FALSE)
@
%
The \pkg{decorr} package training dataset is composed by five stelaes decorations drawings \citep{DiazGuardamino10} belonging to the so-called 'Warrior stelae' family -- with about 140 stelae -- dated to the Late Bronze Age of SW Iberian peninsula \citep{Celestino01}. \\
At first the drawings dataset can be checked by using the \code{imgs} dataframe and the \pkg{magick} package (Fig. \ref{fig:dataset})
%
<<timg, echo=TRUE, message=FALSE, warning=FALSE, fig.pos = '!htbp', fig.align="center", fig.width=15, fig.height=11.2, fig.cap="\\label{fig:dataset}Decoration images of the training dataset">>=
# library("magick")
pth <- system.file("extdata", package = "decorr")
imgs <- read.table(system.file("extdata", "imgs.tsv", package = "decorr"),
                   sep="\t", stringsAsFactors = FALSE)
lims <- list()
for(i in 1:nrow(imgs)){
  i1 <- image_read(paste0(pth,"\\",imgs[i,"img"]))
  lbl.txt <- paste0(imgs[i,"idf"],"\n",
                    imgs[i,"site"],"\n",
                    imgs[i,"decor"],"\n",
                    imgs[i,"img"],"\n",
                    image_info(i1)$width,"*",image_info(i1)$height," px")
  i1 <- image_annotate(i1,lbl.txt,location = "northwest",
                       size = 28, color = "red")
  lims[[length(lims)+1]]<- i1
}
out.img <- image_append(c(image_append(c(lims[[1]],lims[[2]],lims[[3]])),
                          image_append(c(lims[[4]],lims[[5]]))),
                        stack = TRUE)
par(mar=c(0,0,0,0))
plot(out.img)
@
%
To construct a graph overlapping the decoration images listed in the \code{images} dataframe, the first step is to load \code{nodes} and \code{edges} dataframes.
%
<<echo=TRUE>>=
nodes <- read.table(system.file("extdata", "nodes.csv", package = "decorr"),
                    sep=";",stringsAsFactors = FALSE)
edges <- read.table(system.file("extdata", "edges.csv", package = "decorr"),
                    sep=";",stringsAsFactors = FALSE)
@
%
% \begin{itemize}
% \item The \code{nodes} dataframe
% \end{itemize}
\smallbreak
\textbf{\code{nodes} dataframe}
\\
It contains the required minimum variables for the analysis (Tab. \ref{tab:tnds}).

<<tndss, results='asis', echo=TRUE>>=
caption <- "Nodes (from \\code{nodes.csv} dataframe)"
print(xtable::xtable(head(nodes),
                     caption = caption,
                     label="tab:tnds"),
      table.placement="!htbp")
@

Since a prehistorical site can have various decorated objects, the primary key of the decoration is based on two fields: \code{nodes$site} and \code{nodes$decor}. The \code{nodes$id} is the identifier of the node. The \code{nodes$type} field is the default variable for further statistical analysis. Here, \code{nodes$type} refers to the typology of the GUs (anthropomorph, weapons, etc.). The \code{nodes$x} and \code{nodes$y} columns refer to the \emph{x} and \emph{}y coordinates of the nodes. As said, in the first place theses coordinates come from the GIS. But, in a GIS, the coordinates origin (0, 0) is the bottom-left corner and exist negative values, while this origin is top-left for any \proglang{R} matrices (rasters, grids, dataframes, etc.) with only positive values. To recover the local \emph{y} value of GUs nodes and edges on the decoration image, the absolute value \emph{y} value and the image height, as a constant offset, are computed.
\smallbreak
% \begin{itemize}
% \item The \code{edges} dataframe
% \end{itemize}
\textbf{\code{edges} dataframe}
\\
The \code{edges} dataframe is quite similar to the \code{nodes} dataframe (Tab. \ref{tab:teds}).

<<tedss, results='asis', echo=TRUE>>=
caption <- "Edges (from \\code{edges.csv} dataframe)"
print(xtable::xtable(head(edges),
                     caption=caption,
                     label="tab:teds"),
      table.placement="!htbp")
@

Fields \code{edges$site} and \code{edges$decor} are the primary key of decoration. The fields \code{edges$a} and \code{edges$b} are the equivalent to columns \emph{from} and \emph{to} in Graph theory. Even if undirected graphs will the most common models in further studies, this direction helps to distinguish between nodes. The first column \code{edges$a} is the identifier of starting node or \emph{main node} or \emph{overlapping node}. The second column, \code{edges$b}, is the identifier of the ending node or \emph{attribue node} or \emph{overlapped node}. The \code{edges$type} is the type of relation (normal, attribute, overlapping, etc.) between the starting node and the ending node. There is no need to record the coordinates of the edges, these coordinates are calculated from the \code{nodes} dataframe. For example, Table \ref{teds} shows that the Cerro Muriano 1 decoration's first edge connects the nodes \code{1} and \code{8} (respectively in column \code{edges$a} and \code{edges$b}). A way to retrieve these connected nodes' coordinates will be:

<<results='asis', echo=TRUE>>=
cm.1 <- subset(nodes, decor == "Cerro Muriano 1" & id == 1)[,c("x","y")]
cm.8 <- subset(nodes, decor == "Cerro Muriano 1" & id == 8)[,c("x","y")]
paste(as.numeric(cm.1),";",as.numeric(cm.8))
@

Once these dataframes loaded, the list of decoration graphs can be calculated with the \code{list_dec()} function.

\subsection{\code{list\_dec()} function} \label{sec:funlistdec}

The \code{list\_dec()} function allows to calculate graphs for all decorations stored into \code{nodes}, \code{edges} and \code{images}. The result is a list of decoration graph. The first graph of the list can be plotted (Fig. \ref{fig:igraph})

<<igraphs, echo=TRUE, message=FALSE, warning=FALSE, fig.pos = '!htbp', fig.align="center", fig.width=2.5, fig.height=2.5, fig.cap="\\label{fig:igraph}R interface. Plot of the first graph of the list">>=
lgrph <- list_dec(imgs,nodes,edges)
par(mar=c(0,0,0,0))
plot(lgrph[[1]],
     vertex.color = "orange",
     vertex.frame.color="orange",
     vertex.label.color = "black",
     vertex.size = 16,
     vertex.label.cex = 1,
     edge.color = "orange",
     vertex.label.family="Helvetica"
)
@

>>>>

Symbolic practices is a characteristic trait of human societies. Even discussed, such practices seems to start between 233,000 to 800,00 BC \citep{dErrico00}, covering more than 97\% of total human societies time span. Symbolism cover a large range of practices, from ochre deposit in a tomb, to menhir alignements, passing through wall fresco. This latter, what might be called "iconographical practices", probably shows the most complex and interesting testimonies of past societies symbolism. For decades, its study was linked to history of religion because commonly seen as closely linked to cultual practices and believes. Since the \textit{New Archaeology} developpement during the 60's \citep{Clarke14}, symbolic expressions start to be studied with the same formal methods (statistics, seriations, distribution maps, etc.) as any another aspect of social organisation: settlement patterns, tools \emph{chaine op\'eratoire}, susbsitence strategies, etc. \citep{Renfrew91}. But unlike many aspects of the material culture where technological requirements and technical efficiency determine the choice of the raw material and the shape of the object -- a flint blade for cutting, a pottery for containing, a house for living --, the function of an iconographic composition cannot be drawn directly from itself. Furthermore, there is no specific and controled vocabularies to describe hand-made images. There is always a difficulty to decribe an iconographical content with alphanumerical signs. Whether these last decades study of ancient iconography had  undergone significative improvements at the site scale -- with GIS/database statistics, paleoenvironmental restitutions, etc. -- and at the sign scale with the development of archaeological sciences -- radiocarbon dating, use-wear analysis, elemental analysis, etc. --, these improvement do not necessarly help to understand the semantic content of the iconography.\\
Semantics or semiotics can be defined as a system of conventional features -- called signs -- organised also in conventional manners. Until our days, formal methods to study ancient iconography semantics have been mostlty been grounded  -- explicitly or not -- on the prime principle of Saussurian linguistic: the '\emph{linearity of the signifier}' \citep{Saussure89}.
Writing is one of the most rational semiographical system with a clear distinction between signified and signifier and the development of the signified on a linear axis. Even if we do not understand the meaning of the signs, writing can easily be modelled with Graph theory and recurent patterns can be identified. For example, the 3-letters word "\code{art}" can be modelled with three vertices (\code{a}, \code{r}, \code{t}) and two edges (one between \code{a} and \code{r}, the other between \code{r} and \code{t}). In \proglang{R}, these features, concatenated in this order with a \code{paste0()}, is \code{art}, and not \code{rat} (Fig. \ref{fig:fart}).

@@

Because of the inherent variability of iconography, and because graphical and spatial proximities between graphical units are generaly not quantified, applying the Saussurian model to any prehistorical graphical content had led to considerable problems:
\begin{itemize}
\setlength\itemsep{.05em}
\item unexplicit spatial groupings of graphical units -- like graphical units grouped into \emph{figures},  \emph{figures} grouped into \emph{patterns}, \emph{patterns} grouped into \emph{motives}, etc. -- with tedious number of groups
\item consistency, proximities and relationships between these groups are often implicit and not quantified
\item studies develop proper descriptive vocabularies, singular relationships of grouping, idosyncratic methods at site-dependent or period-dependent scales
\end{itemize}

Furthermore, at the graphical units level, top-down hierachical typologies (e.g., a category of graphical units with \emph{n} types, a type with \emph{n} subtypes, a subtype with \emph{n} varieties, etc.) lead also to tedious inventories and difficult the comparisons \emph{within} and \emph{between} the categories.

@@

According to us, prehistorical composition -- like any formal system -- are spatial features (mostly shapes) related one with the other depending on rules of spatial proximities. Then, the First Law of Geography: "\emph{everything is related to everything else, but near things are more related than distant things}" \citep{Tobler70} should be more operative than the Saussurian '\emph{linearity of the signifier}' principle \citep{Saussure89}. To get these '\emph{near things}', or features proximities, we use concepts coming from the Graph Theory, which offers a general framework to manage neighbourhoods, and particularly relative neighborhood graphs (RNG) which is '\emph{a powerful model of low-level visual process}' \citep{Toussaint80}. Furthermore, we address the questions of managing graphical units attributes (for example an anthropomorph with a sword \emph{vs} an anthropomorph with a spear) and superimpositions of graphical units.

@@

Graph elements, nodes and edges -- repectively GUs and connexions between GUs -- are created on a GIS interface. Indeed, for large series of decorated supports, GIS offers the most suitable and flexible interface to register all GUs and to get their coordinates. The decoration image is considerated as the basemap of the project and will cover the region of interest of the analysis (ie, all the graphical content). The decoration image can be binarized: GUs are considerated active, the undecorated parts of the support -- the background -- are considerated inactive. The polygonisation of the GUs (\code{POLYGONS}), including the border of the stelae, allows to calculated their Voronoi seeds (ie, their centroid) and Voronoi cells (ie, their area of influence). So, the entire decoration image is tiled and condidered as a Voronoi diagram. Centroids of the GUs (\code{POINTS}) are calculated and considered as the graph nodes, their locations (\emph{x} and \emph{y} coordinates, measured in pixels) are relative to each decoration. Exist edges (\code{LINES}) between pairs of GUs when their Voronoi cells share a common border (\emph{birel}: 'touches').

The entire graphical content of the decoration is mapped with a neighborhood graphs (RNG) -- ie geometric graph, or planar graph -- of GUs centroids (nodes) and GUs proximity links (edges). So, a decoration graph is a one component graph. This graph is the dual graph of the decoration Voronoi diagram (Fig. \ref{fig:gis}).

\begin{figure}[H]
% \begin{figure}[!htbp]
\centering
\includegraphics{gis}
\caption{\label{fig:gis} GIS interface. A) Original decoration of the Late Bronze Age Cerro Muriano 1 stelae (drawing: \cite{DiazGuardamino10}) opened in new GIS project without any projection system with its extent (\code{xmin, xmax, ymin, ymax} in pixels); B) Polygonisation of the GUs, calculation of their centroids (nodes), calculation of the Voronoi cells, calculation of the proximity links (edges); C) plot of the relative \emph{x} and \emph{y} coordinates of the GUs centroids; D) plot of the values in the \code{type} of the GUs}
\end{figure}

Whether this theoritical process include image binarization, polygonization and \emph{birel} analysis, a simplier solution will be to create directly centroids on the GUs and to draw the edges manually. Then \emph{x} and \emph{y} coordinates of nodes can be easily retrieved.\\

\subsection[elements]{Graph elements} \label{sec:elements}

This model has a minimal of \emph{a priori} definitions. Those definitions only concern the graph elements: nodes variables (ie, GUs variables) and types of relations they share (edges).

\smallbreak
\textbf{Nodes types}
\\
At least one variable, like the \code{type} of the nodes, is defined in order to analyse and compare iconographical compositions. In this article, we will only consider nodes \code{type} values ('anthropomorph', 'sword', etc.) but it the package user can add other variables (\code{technique}, \code{color}, etc.)

\smallbreak
\textbf{Egdes types}
\\
The package considers the most common types of relations between nodes:

\begin{description}
\item[\emph{normal edges}]
Two contemporaneous nodes having a Voronoi cell sharing a border have a common edge. By convention, this edge is tagged \code{'='} (\code{type = '='}), represented with an orange plain line and its textual notation is \code{'-=-'}. For example: \code{personnage -=- bouclier} means that the nodes \code{personnage} and \code{bouclier} have a common border (Fig. \ref{fig:gis_1}). As said, in Prehistorical iconography, the read direction is not known, so this edge is undirected, ie \code{personnage -=- bouclier} is equal to \code{bouclier -=- personnage} (Fig. \ref{fig:gis_1}).

\item[\emph{attribute edges}]
It occurs frequently that a node can be divided into a main unit (eg, an anthropomorph) and one or various attribute units (eg, a helmet, a male sex). To record this ordered set, the edge between a main and an attribute node is tagged \code{'+'} (\code{type = '+'}), represented with an orange dashed line and its textual notation is \code{'-+-'}. For example \code{personnage -+- casque} means that the main node \code{personnage} has the attribute node \code{casque}. The edge is directed since \code{personnage -+- casque} (the anthropomorph has an helmet) is not equal to \code{casque -+- personnage} (the helmet has an anthropomorph) (Fig. \ref{fig:gis_1}).

\begin{figure}[H]
% \begin{figure}[!htbp]
\centering\emph{}
\includegraphics[height=10cm, trim=15mm 20mm 15mm 20mm, clip, keepaspectratio]{gis_1}
\caption{\label{fig:gis_1} GIS interface, Cerro Muriano 1 decoration. The nodes \code{casque} (helmet) and \code{sexe\_masculin} (male sex) are two nodes attributes of the node \code{personnage} (anthropomorph).}
\end{figure}

Broadly speaking, for further statistical analysis, it is better to use this \emph{attribute} method than to multiply the categories of nodes.

\item[\emph{overlapping edges}]
Finally, it occurs that a graphical composition shows decoration phases. These decoration phases can be identified by observing the superimpositions between different nodes or by other means (the technique or the patina for example). In the first case, two nodes belonging to two different phase will have a common part (the overlapped intersection) while in the second case the two nodes will not necessarly share a part of their surface. To record this relative chronology, an edge between two non-contemporaneous nodes is tagged \code{'>'} (\code{type = '>'}), represented with a blue plain line and its textual notation is \code{'->-'}. For example \code{ecriture ->- bouclier} means that \code{ecriture} is more recent than the \code{bouclier}, or \emph{overlaps} the\code{bouclier} in the stratigraphical sense. The edge is directed since \code{ecriture ->- bouclier} (ie, \code{ecriture} \emph{overlaps} \code{bouclier} ) is not equal to \code{bouclier ->- ecriture} (ie, \code{bouclier} \emph{overlaps} \code{ecriture}) (Fig. \ref{fig:gis_2}).

@@

\begin{figure}[H]
% \begin{figure}[!htbp]
\centering
\includegraphics[height=10cm, trim=30mm 16mm 30mm 13mm, clip, keepaspectratio]{gis_2}
\caption{\label{fig:gis_2} GIS interface, Ibahernando decoration. The node \code{ecriture} (writing) overlaps \code{lance} (spear) and the upper part of the \code{bouclier} (shield).}
\end{figure}

This stratigraphical information (node 1 \emph{over} node 2, or node 2 \emph{under} node 1) helps to understand the relative chronology between node and must be recorded. The analysis of the iconographical content can be stratified (stratified analysis) and performed on each different phases of decoration separately.

@@

\begin{table}[!htbp]
\centering
\begin{tabular}{ccccp{2cm}p{4.9cm}}
\hline
% node A & type of edge & node B & undirected & birel & stratigraphical meaning \\
Node 1 & Edge type & Node 2 & Edge directness & \emph{Birel} & Description \rule{0pt}{1em} \\[1pt]
\hline \rule{0pt}{1em}%
\code{A} & \code{=} & \code{B} & Undirected & $ A \cap B = \emptyset $ & 
  \code{A} and \code{B} are contemporaneous and disjoint. \\
\code{A} & \code{+} & \code{B} & Directed & $ A \cap B = A $ & 
  \code{A} and \code{B} are contemporaneous. \code{B} is an attribute of \code{A}. \\
\code{A} & \code{>} & \code{B} & Directed & $ A \cap B = \exists $ \newline or \newline $ A \cap B = \emptyset $ & 
  \code{A} is more recent than \code{B}. \\
% \code{1} & = & 2 & undirected & A \cap B = \varnothing & 1 and 2 can be contemporaneous \\
\hline
\end{tabular}
\caption{Synthesis for the different types of relations between GUs.}\label{tab1}
\label{table:tedges}
\end{table}

@@

A training dataset (nodes and edges coordinates, decoration images) is stored in the \code{extdata} folder of the \pkg{decorr} package
\smallbreak
% \begin{itemize}
% \item The \code{imgs} dataframe
% \end{itemize}
\textbf{\code{imgs} dataframe}
\\
The inventory of decorations is stored in the \code{imgs} dataframe (Tab. \ref{tab:timg}). The field \code{imgs$idf} is the short name of the decoration, useful during statistical analysis. The primary key of each decoration is the concatenate of \code{imgs$site} and \code{imgs$decor}. These keys will allow joints with the other dataframes (\code{nodes} and \code{edges})
%
<<results='asis', echo=FALSE, warning=FALSE>>=
imgs <- read.table(system.file("extdata", "imgs.tsv", package = "decorr"),
                   sep="\t", stringsAsFactors = FALSE)
print(xtable::xtable(imgs,
    caption="The studied corpus, the \\code{imgs.tsv} dataframe",
    label="tab:timg",
    size=6,
    align=c(
      "p{0.01cm}",
      "c",
      ">{\\raggedright\\arraybackslash\\rule{0pt}{1em}\\addtolength{\\baselineskip}{-1.5pt}}p{2.1cm}",
      ">{\\raggedright\\arraybackslash\\addtolength{\\baselineskip}{-1.5pt}}p{2.1cm}",
      "p{8.5cm}")),
    table.placement="!htbp",
    include.rownames=FALSE)
@
%
The \pkg{decorr} package training dataset is composed by five stelaes decorations drawings \citep{DiazGuardamino10} belonging to the so-called 'Warrior stelae' family -- with about 140 stelae -- dated to the Late Bronze Age of SW Iberian peninsula \citep{Celestino01}. \\
At first the drawings dataset can be checked by using the \code{imgs} dataframe and the \pkg{magick} package (Fig. \ref{fig:dataset})
%
<<timg, echo=TRUE, message=FALSE, warning=FALSE, fig.pos = '!htbp', fig.align="center", fig.width=15, fig.height=11.2, fig.cap="\\label{fig:dataset}Decoration images of the training dataset">>=
pth <- system.file("extdata", package = "decorr")
imgs <- read.table(system.file("extdata", "imgs.tsv", package = "decorr"),
                   sep="\t", stringsAsFactors = FALSE)
lims <- list()
for(i in 1:nrow(imgs)){
  i1 <- image_read(paste0(pth,"\\",imgs[i,"img"]))
  lbl.txt <- paste0(imgs[i,"idf"],"\n",
                    imgs[i,"site"],"\n",
                    imgs[i,"decor"],"\n",
                    imgs[i,"img"],"\n",
                    image_info(i1)$width,"*",image_info(i1)$height," px")
  i1 <- image_annotate(i1,lbl.txt,location = "northwest",
                       size = 28, color = "red")
  lims[[length(lims)+1]]<- i1
}
out.img <- image_append(c(image_append(c(lims[[1]],lims[[2]],lims[[3]])),
                          image_append(c(lims[[4]],lims[[5]]))),
                        stack = TRUE)
par(mar=c(0,0,0,0))
plot(out.img)
@
%
To construct a graph overlapping the decoration images listed in the \code{images} dataframe, the first step is to load \code{nodes} and \code{edges} dataframes.
% \begin{itemize}
% \item The \code{nodes} dataframe
% \end{itemize}
\smallbreak
\textbf{Nodes dataframe}
\\
It contains the required minimum variables for the analysis (Tab. \ref{tab:tnds}).

<<tndss, results='asis', echo=TRUE>>=
caption <- "Nodes (from \\code{nodes.csv} dataframe)"
nodes <- read.table(system.file("extdata", "nodes.csv", package = "decorr"),
                    sep=";",stringsAsFactors = FALSE)
print(xtable::xtable(head(nodes),
                     caption = caption,
                     label="tab:tnds"),
      table.placement="!htbp")
@

Since a prehistorical site can have various decorated objects, the primary key of the decoration is based on two fields: \code{nodes$site} and \code{nodes$decor}. The \code{nodes$id} is the identifier of the node. The \code{nodes$type} field is the default variable for further statistical analysis. Here, \code{nodes$type} refers to the typology of the nodes (anthropomorph, weapons, etc.). The \code{nodes$x} and \code{nodes$y} columns refer to the \emph{x} and \emph{}y coordinates of the nodes. As said, in the first place theses coordinates come from the GIS. But, in a GIS, the coordinates origin (0, 0) is the bottom-left corner and exist negative values, while this origin is top-left for any \proglang{R} matrices (rasters, grids, dataframes, etc.) with only positive values. To recover the local \emph{y} value of nodes and edges on the decoration image, the absolute value \emph{y} value and the image height, as a constant offset, are computed.
\smallbreak
% \begin{itemize}
% \item The \code{edges} dataframe
% \end{itemize}
\textbf{\code{edges} dataframe}
\\
The \code{edges} dataframe is quite similar to the \code{nodes} dataframe (Table \ref{tab:teds}).

<<tedss, results='asis', echo=TRUE>>=
caption <- "Edges of Cerro Muriano 1 (from \\code{edges.csv} dataframe)"
edges <- read.table(system.file("extdata", "edges.csv", package = "decorr"),
                    sep=";",stringsAsFactors = FALSE)
edges.cm1 <- subset(edges, decor == "Cerro Muriano 1")
print(xtable::xtable(head(edges.cm1),
                     caption=caption,
                     label="tab:teds"),
      table.placement="!htbp")
@

Fields \code{edges$site} and \code{edges$decor} are the primary key of decoration. The fields \code{edges$a} and \code{edges$b} are the equivalent to columns \emph{from} and \emph{to} in Graph theory. Even if undirected graphs will the most common in further studies, this direction helps to distinguish between nodes. The column \code{edges$a} is the identifier of \emph{starting node} or \emph{main node} or \emph{overlapping node}. The column \code{edges$b} is the identifier of the \emph{ending node} or \emph{attribue node} or \emph{overlapped node}. The column \code{edges$type} is the type of relation (normal, attribute, overlapping, etc.) between the node a and the node b. There is no need to record the coordinates of the edges, these coordinates are calculated from the \code{nodes} dataframe.

The \pkg{decorr} package functions can be divided into:
\begin{enumerate}
\item management functions
\item graphical functions
\item single decoration functions
\item comparisons between different decorations functions
\end{enumerate}

Once these dataframes loaded, the list of decoration graphs can be calculated with \code{list_dec()}.

\subsection{Management functions} \label{sec:manage}

\end{Begin}
list_dec(imgs,
         nodes,
         edges)
\end{Code}

Once \code{imgs}, \code{nodes} and \code{edges} dataframes loaded, the \code{list\_dec()} function allows to create graphs for all decorations within these dataframes. The result is a list of \code{igraph} objects. As an example, the first graph of the list can be plotted with the \code{igraph::plot()} function (Fig. \ref{fig:igraph})

<<igraphs, echo=TRUE, message=FALSE, warning=FALSE, fig.pos = '!htbp', fig.align="center", fig.width=2.5, fig.height=2.5, fig.cap="\\label{fig:igraph}R interface. Plot of the first graph of the list (Cerro Muriano 1 decoration)">>=
lgrph <- list_dec(imgs, nodes, edges)
par(mar=c(0, 0, 0, 0))
plot(lgrph[[1]],
     vertex.color = "orange",
     vertex.frame.color = "orange",
     vertex.label.color = "black",
     vertex.size = 16,
     vertex.label.cex = 1,
     edge.color = "orange",
     vertex.label.family = "Helvetica"
)
@

\begin{Code}
named_elements(grph,
               focus = "edges",
               nd.var = "type")
\end{Code}

The \code{named\_elements()} function allows to calculate the textual expression of graph elements (nodes and edges) with a disambiguation of these elements -- adding '#' suffixes -- when several nodes or edges have the same values.
